---
title: "DataExploration"
output: html_document
date: "2024-10-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

base_path <- "C:/Users/oaw001/Documents/AnomalyDetection"
scripts <-
  list(
    "LocalInitialisationScript.R",
    "DataExploration.R"
  )
source_script <- function(script) {
    source(file.path(base_path, "Scripts", script))
}
walk(scripts, source_script)

```

## Data Exploration

This is the R markdown for exploring the data for the one-class-classification workflow after the test data has been removed. This document will plot the behavioural examples, the volumes of data, as well as perform analysis to determine the optimal window durations for each behaviour.

All plot functions are provided at the bottom of the Rmd file (hidden in pdf output) and can be edited. While this file can be run independently, it is easier to run in sequence with the other scripts as I call on information and data from the main initialisation script.

### Plot examples of the data
The following plots will show trace examples from each category and the first 5 individuals. 5 individuals were selected to limit the size of the plot, but this number be changed in the Rmd file.

This plot can be used to assess the consistency of the behavioural annotations. If there are any behaviours that look very similar to each other, or dissimilar from each other, classification may be difficult, and more exploration may be necessary.

```{r plot_trace_examples, echo=FALSE}
  plot_trace_examples <- plotTraceExamples(behaviours = unique(data_other$Activity), 
                                         data = data_other, 
                                         individuals = 9,
                                         n_samples = 250, 
                                         n_col = 4)
  plot_trace_examples
```

### Plot volumes of data
This plot shows the volume of data for each behavioural class in the non-test dataset. It helps in assessing class imbalance, noting that categories with fewer samples may be harder to classify, as well as relative minority classes struggling compared to majority classes. Additionally, the plot shows how individuals are distributed across different activities. Note whether there are any behaviours that are not well distributed across multiple individuals. If so, this may cause problems when not every training/validation split contains the same behaviours.

```{r plot_activity_by_ID, echo=FALSE}
  plot_activity_by_ID <- plotActivityByID(data = data_other, 
                                        frequency = sample_rate, 
                                        colours = length(unique(data_other$ID)))
  plot_activity_by_ID
```

### Behaviour durations
In order to select the custom window length for each behaviour's OCC model, the duration of the behaviours can be assessed. A boxplot and summary statistics can be used to explore the averages and spreads of the bout durations (where bouts are continuous sequences of the same behavioural type).
```{r plot_behaviour_durations, echo=FALSE}

  plot_behaviour_durations <- BehaviourDuration(data = data_other, 
                                              sample_rate = sample_rate, 
                                              target_activities = unique(data_other$Activity))
  
  plot <- plot_behaviour_durations$duration_plot
  plot
  stats <- plot_behaviour_durations$duration_stats
  ktable(stats)
```


```{r Functions, include= FALSE}
plotTraceExamples <- function(behaviours, data, individuals, n_samples, n_col) {
  
  data <- data %>% filter(ID %in% sample(unique(data$ID), individuals))
  
  # Create plots for each behavior (with error catching)
  plots <- purrr::map(behaviours, function(behaviour) {
    tryCatch(
      {
        plot_behaviour(behaviour, n_samples, data)
      },
      error = function(e) {
        message("Skipping plot for ", behaviour, ": ", e$message)
        NULL  # Return NULL to indicate skipping
      }
    )
  })
  
  # Remove NULL plots (for behaviors with no data)
  plots <- purrr::compact(plots)
  
  # Combine plots into a single grid
  grid_plot <- cowplot::plot_grid(plotlist = plots, ncol = n_col)
  
  return(grid_plot)
}

# Function to create the plot for each behavior
plot_behaviour <- function(behaviour, n_samples, data) {
  df <- data %>%
    filter(Activity == behaviour) %>%
    group_by(ID, Activity) %>%
    slice(1:n_samples) %>%
    mutate(relative_time = row_number())
  
  # Check if the filtered dataframe is empty
  if (nrow(df) == 0) {
    stop("No data available for behaviour: ", behaviour)
  }
  
  ggplot(df, aes(x = relative_time)) +
    geom_line(aes(y = Accelerometer.X, color = "X"), show.legend = FALSE) +
    geom_line(aes(y = Accelerometer.Y, color = "Y"), show.legend = FALSE) +
    geom_line(aes(y = Accelerometer.Z, color = "Z"), show.legend = FALSE) +
    labs(title = paste(behaviour),
         x = NULL, y = NULL) +
    scale_color_manual(values = c(X = "salmon", Y = "turquoise", Z = "darkblue"), guide = "none") +
    facet_wrap(~ ID, nrow = 1, scales = "free_x") +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank())
}

## set colours ####
generate_random_colors <- function(n) {
  colors <- rgb(runif(n), runif(n), runif(n))
  return(colors)
}

## total volume by Activity and ID ####
plotActivityByID <- function(data, frequency, colours) {
  my_colours <- generate_random_colors(colours)
  # summarise into a table
  labelledDataSummary <- data %>%
    #filter(!Activity %in% ignore_behaviours) %>%
    count(ID, Activity)
  
  # account for the HZ, convert to minutes
  labelledDataSummaryplot <- labelledDataSummary %>%
    mutate(minutes = (n/frequency)/60)
  
  # Plot the stacked bar graph
  plot_activity_by_ID <- ggplot(labelledDataSummaryplot, aes(x = Activity, y = minutes, fill = as.factor(ID))) +
    geom_bar(stat = "identity") +
    labs(x = "Activity",
         y = "minutes") +
    theme_minimal() +
    scale_fill_manual(values = my_colours) +
    theme(axis.line = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1),
          panel.border = element_rect(color = "black", fill = NA),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
  
  return(plot_activity_by_ID)
}

## duration of each of the behaviours ####
BehaviourDuration <- function(data, sample_rate, target_activities){
  
  summary <- data %>%
    mutate(Activity = ifelse(Activity %in% target_activities, Activity, "Other")) %>%
    arrange(ID) %>%                     # Sort by ID (not time because multiple trials in dog data)
    group_by(ID) %>%                    # Group by ID
    mutate(
      behavior_change = lag(Activity) != Activity,  # Detect changes in Activity
      behavior_change = ifelse(is.na(behavior_change), TRUE, behavior_change)  # Handle the first row
    ) %>%
    mutate(
      behavior_id = cumsum(behavior_change)  # Create an identifier for each continuous behavior segment
    ) %>%
    group_by(ID, behavior_id) %>%            # Group by ID and behavior_id
    mutate(
      row_count = row_number()                # Count rows within each behavior segment
    ) %>%
    ungroup() %>%
    select(ID, Time, Activity, row_count, behavior_id) %>%   # Select relevant columns
    group_by(ID, Activity, behavior_id) %>%
    summarise(duration_sec = max(row_count)/100) %>%
    filter(Activity %in% target_activities)
  
  duration_stats <- summary %>%
    group_by(Activity) %>%
    summarise(
      average_behaviour_duration = mean(duration_sec, na.rm = TRUE),
      median_behaviour_duration = median(duration_sec, na.rm = TRUE),
      variation_duration = sd(duration_sec, na.rm = TRUE),
      maximum_duration = max(duration_sec, na.rm = TRUE),
      minimum_duration = min(duration_sec, na.rm = TRUE),
      third_quartile_duration = quantile(duration_sec, 0.25, na.rm = TRUE),  # Add third quartile
      count = n()
    )
  
  # plot that
  duration_plot <- ggplot(summary, aes(x = Activity, y = as.numeric(duration_sec))) +
    geom_boxplot(aes(color = Activity)) +  # Use color to distinguish activities
    theme_minimal() +
    theme(
      legend.position = "none",             # Remove legend
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),  # Rotate x-axis labels 90 degrees
      panel.grid = element_blank(),         # Remove grid lines
      panel.border = element_rect(color = "black", fill = NA)  # Add black border around the plot
    ) +
    labs(
      x = "Activity",
      y = "Duration (seconds)"
    ) +
    scale_y_continuous(
      limits = c(min(summary$duration_sec, na.rm = TRUE), max(summary$duration_sec, na.rm = TRUE)),  # Set y-axis limits
      breaks = seq(0, max(summary$duration_sec, na.rm = TRUE), by = 160)  # Adjust the step size as needed
    )
  
  return(list(duration_plot = duration_plot,
              duration_stats = duration_stats))
}
```

